---
title: "Machine Learning - Homework 10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


### ----------------------------------------------------------------------------

# Instructions

**This homework will be graded on completion of all questions. Peer reviews are not required.** 


# Objectives

* Practice with PCA

* Practice with clustering methods


### ----------------------------------------------------------------------------

For questions 1-3, we will work with the `NCI60` microarray dataset. The dataset contains expression levels on 6830 genes from 64 cancer cell lines.

```{r}
library(ISLR2)   # load package

ncidata <- NCI60$data   # load dataset

dim(ncidata)    # dimension of the dataset (number of rows x number of columns)
```


### ----------------------------------------------------------------------------

## Question 1

(a) Implement PCA on the standardized (centered and scaled) data. How many PCs are required to explain at least 75% of the overall variance?

```{r}
# write your R code here
set.seed(10)



```

```{}
# write your answer here



```


(b) Create a scree plot showing all the possible number of PCs. Suggest a suitable number of PCs for this data by examining the scree plot. [See help page of `screeplot` function to specify the number of components to be plotted]

```{r}
# write your R code here
set.seed(10)



```

```{}
# write your answer here



```


### ----------------------------------------------------------------------------

## Question 2

Implement K-means clustering on the standardized (centered and scaled) `NCI60` microarray data with $K=4$ and 20 random initial configurations. What are the cluster sizes?

```{r}
# write your R code here
set.seed(10)



```

```{}
# write your answer here



```


### ----------------------------------------------------------------------------

## Question 3

(a) Implement hierarchical clustering on the standardized (centered and scaled) `NCI60` microarray data using complete, single, and average linkages. Comment on how the dendrograms from the different linkage methods compare. [See help page of `hclust` function to specify the linkage method.]

```{r}
# write your R code here
set.seed(10)



```

```{}
# write your answer here



```


(b) Cut the dendrogram built with complete linkage to have 4 clusters. What are the cluster sizes?

```{r}
# write your R code here
set.seed(10)



```

```{}
# write your answer here



```


(c) For this data, what are the minimum and maximum number of clusters `k` one can specify in the `cutree` function? What will be the corresponding cluster sizes? Explain.

```{}
# write your answer here



```


### ----------------------------------------------------------------------------

## Question 4

Suppose you have a standardized data matrix $X$ with $n$ observations and $p$ features. What will be the total within-cluster variation (total within-cluster sum of squares) when you implement K-means with $K = n$? What will be the corresponding cluster sizes? Explain your answer.

```{}
# write your answer here



```


### ----------------------------------------------------------------------------
