---
title: "Machine Learning - Quiz 4"
author: "Your name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
library(recipes)
library(rpart)
library(ipred)
library(e1071)
library(ranger)
library(kernlab)
```

### ----------------------------------------------------------------------------

## Question 1  

We will work with the `titanic.rds` dataset which contains information on Titanic passengers. Please download the dataset from Canvas, upload it to Posit Cloud, and load it using the following code.

```{r}
titanic <- readRDS("titanic.rds")    # load dataset
```


The variables in the dataset are

* `pclass` (passenger ticket class - '1st', '2nd', '3rd')
* `survived` ('Yes' or 'No')
* `sex` ('male' or 'female')
* `age` (passenger age in years)
* `sibsp` (number of siblings or spouses aboard)
* `parch` (number of parents or children aboard)


The task is to predict `survived` using the rest of the variables in the dataset. Compare the performance (in terms of **Accuracy**) of the following models:

* a single classification tree;

* a random forest model;

* an SVM with linear kernel (support vector classifier), and

* an SVM with radial kernel.


**Perform the necessary tasks.**

* Use a 80-20 split.

* Use 5-fold CV with 1 repeat.

* Parameter grids are given below.

```{r}
# use tuneLength = 20 for the single tree

# for random forest
param_grid_rf <- expand.grid(mtry = seq(1, 5, 1),    
                          splitrule = "gini",
                          min.node.size = 2)


# for SVM with linear kernel 
param_grid_linear <- expand.grid(C = c(0.001, 0.1, 1, 5, 10, 100))    


# for SVM with radial kernel
param_grid_radial <- expand.grid(sigma = c(0.5, 1, 1.5, 2),
                                 C = c(0.001, 0.1, 1, 5, 10, 100))    

```


**Create the confusion matrix with the test set and report the test set accuracy**.

### ----------------------------------------------------------------------------

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```

```{r}
set.seed(052824)


```
### ----------------------------------------------------------------------------
